Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-04-05 LaTeX/response.tex

The turnover within a species from one genetic means of sex determination to another is surprisingly common, and it poses a challenging puzzle for evolutionary geneticists.  Existing theory on the topic suggests that there must be either sexually-antagonistic selection or selection variants that bring the sex ratio closer to 1:1.  This paper builds explicit models of the situation and asks about the role of selection in the haploid stage (i.e. meiotic drive and gametic selection), and finds that such effects can also serve as the primary driver of turnover of sex determining mechanisms.  While this is a rather focused topic, it is a striking evolutionary puzzle, and the paper presents a very satisfying solution that is much more general than we had previously thought.  It also overturns the primacy of Fisherian sex ratio evolution as the driver of the dynamics of genes of this sort.

The authors devise a satisfyingly simple genetic scenario, with one ancestral sex-determining locus, one locus under selection, and a third locus for the nascent sex determining locus.  They manage to collapse some messy algebra into terms like the ``haplotypic growth rate" that allow the reader to intuit many of the results.  I only have a few suggestions:

\noindent\subsection{}
The selected locus allows for three modes of selection: gametic selection (weighting the frequencies of gamete genotypes), the usual diploid viability selection, and meiotic drive (where heterozygotes may produce gametes that deviate from 50:50 proportions).  It should be noted that the model does not accommodate fertility or sexual selection.

\noindent\paragraph{Response}
\textcolor{red}{One could interpret our diploid selection as fertility selection, or am I missing something? Sexual selection does seem like a different model though, with for example, male fitnesses depending on female frequencies; a note such as "We do not consider sexual selection" in the paragraph beginning near L111 might do it?}

@@

\noindent\subsection{}
Gametic selection among sperm is a bit different from sperm competition, since the latter involves pair-wise (or multi-way) bouts, whereas gametic selection assigns constant fitnesses to each haplotype. I would replace occurrences of ``sperm competition" with ``male gametic selection", or maybe ``sperm gametic competition". To be consistent, replace ``egg competition" with ``female/egg gametic competition".

\noindent\paragraph{Response}
\textcolor{blue}{We have made the suggested replacements.
%n the introduction we mention that sperm competition has been shown to alter allele frequencies and fitness, and in this case we now mention this was shown within a single ejaculate (L68).
}

\noindent\subsection{}
I am not certain of this, but it seems that a potentially important difference between ESD and GSD has to do with the variance in proportion male? Since ESD (in a constant environment, as is implicitly modeled here) has a binomial sampling of sexes, but GSD might deviate from this.  This might be an issue because Fisherian sex ratio selection acts on the mean and not the variance.

\noindent\paragraph{Response}
\textcolor{blue}{In our deterministic model (where the number of binomial trials is infinite), the variance in the fraction male is 0 with ESD, as it is with GSD. 
We now mention this on L392: ``In our deterministic model this means the fraction female is exactly $k$, even when $m$ is rare (exploring the effect of the variance induced by ESD would also be interesting)."
 }

@@

On the other hand, I am much less sure that this paper will reach beyond experts in this subject area, though the authors should seek to broaden the appeal. This is not because the results are narrow -- I think they can have a broad appeal. Rather, the paper is written in a technical manner that is likely to limit the appeal beyond the expert class. I am reasonably comfortable with models like this, but still found the presentation to be very hard going in a variety of places, particularly in the results. I offer some suggestions for presentation that may help improve the presentation and broaden the potential appeal of the paper.

\noindent\subsection{}
I think the introduction could be simplified so that the context of the study is more obvious. The authors essentially need to make three key points, which are there, but could be better drawn out. First, transitions between sex-determining systems are common, and recent empirical work has given us a much greater appreciation for just how common these transitions are. Second, there are two major mechanisms that are thought induce evolution of new sex-determination mechanisms: (1) selection to stabilize the sex ratio, and (2) hitchhiking between (tightly linked) sex-determination and sexually antagonistic alleles. Third, the authors should clearly outline the major limitations of current theory as a means of motivating their mega-model that captures an array of genetic and selection scenarios that may lead to transitions in sex determination. As I mentioned, these elements can all be found in the introduction, but the pitch can be substantially streamlined to allow readers to easily see: (1) what is the context? (2) What do we currently think drives transitions between SD systems? And (3) what are the holes in the theory and how will the authors address them?

\noindent\paragraph{Response}
\textcolor{blue}{
The introduction has been simplified to be as follows.
The first paragraph discusses how common transitions are (context). 
The second and third paragraphs discuss the two major mechanisms currently hypothesized to drive the observed transitions, respectively (current theory). 
The fourth paragraph explains how assuming ancestrally-tight sex-linkage produces previously unexpected transitions (one hole in current theory, and how we address it).
The fifth paragraph discusses the potential complications of haploid selection for transitions (a second hole in current theory).
The sixth paragraph describes previous studies on transitions with haploid selection (attempts to address the second hole). 
The final, seventh, paragraph states how our approach differs from previous attempts and the major findings we uncover (how we address the second hole).  
}

@@

\noindent\subsection{}
Likewise, the results yield three major insights:\\
(1) SD transitions can, counterintuitively, potentially INCREASE sex ratio bias\\
(2) Tight linkage and sexual antagonism are not required for transitions to occur\\
(3) Under haploid selection, conditions for the spread of new sex-determining alleles are extremely permissive.\\
Again, all of these points are there, but they could be better drawn out than they currently are. The results, as written, are hard-going, and I expect will be particularly challenging for readers that are not heavily immersed in evolutionary theory.

One potential way to help readers -- particularly those interested in what the paper has to say, but are not mathematically inclined -- is to conspicuously highlight each important result and briefly place it into context of earlier theory. But at the same time, the authors have done a great job of drawing out general mathematical results from a relatively complicated model -- their mathematical insights ARE the results and I am not suggesting that they be minimized or shunted into appendices.

One tactic for including analytical details which specialists will want to see, while allowing less mathematically inclined readers to easily find each main result, is to place short and accessible summaries at the close of each section of the Results. Such a tactic has been put to good use in a hugely influential paper by Allen Orr (1998 Evolution), which I expect the authors will be familiar with. Orr presents the major mathematical results of his analysis alongside prominently labelled punchline paragraphs that summarize each of his main results in a concise and easily understandable manner. If the authors choose to take this approach, please be clear to place the take-home messages in context: clearly state whether the result confirms previous theory, or represents a new result or contradiction of an earlier result.

\noindent\paragraph{Response}
\textcolor{blue}{
%We now give 3 main conclusions at the end of the introduction (L69) which echo those mentioned by the reviewer (looser linkage can evolve, equal selection to increase or decrease the sex ratio, ESD does not always invade biased sex ratio; all of which suggest haploid selection makes conditions for transitions more permissive).
We now give 4 main conclusions in the results, in the style of Orr 1998:\\
%(1) transitions are determined by both sex-ratio selection and selection on associated alleles,\\
%(2) cis- and trans-GSD transitions are qualitatively different,\\
(1A) ancestrally-tight sex-linkage allows looser sex-linkage to evolve in trans-GSD transitions (with or without haploid selection),\\
(1B) haploid selection allows looser sex-linkage to evolve in trans-GSD transitions,\\
(2) transitions can occur when new sex-determining alleles increase linkage with haploid selected loci,\\
(3) when selection is weak relative to recombination, sex-ratio selection is exactly equal in magnitude to the selection imposed by associations with selected alleles, and\\
(4) transitions from genetic to environmental sex-determination are not straightforwardly predicted by selection to balance the zygotic sex ratio when haploid selection is present.
\vspace{0.25cm}\\
As (1A), (1B), and (2) have some antecedents in the literature, we discuss these immediately below each conclusion. 
%We also discuss, immediately below each conclusion, its novelty.
%These are all new results, although (1A) was shown for the special case of $r=0$, $R=1/2$, and ploidally-antagonistic selection in Kozielska et al 2010, which we now mention immediately below (1A). 
}

@@

\noindent\subsubsection{}
L56-57: Rephrase for clarity "individual cases of meiotic drive are generally sex-limited, and exclusively effect male or female gametes". And on line 57, "sex specific" could be changed to "sex-limited".

\noindent\paragraph{Response}
\textcolor{blue}{Rewritten as suggested.}

\noindent\subsubsection{}
Lines 80-81: This primary finding is completely aligned with the earlier views of theory that are summarized above. Are you sure you want to lead with this as the main emphasis of the study? It understates what is new.

\noindent\paragraph{Response}
\textcolor{blue}{We have rewritten the introduction, as well as our main conclusions (see Responses 2.1 and 2.2).}

@@

\noindent\subsubsection{}
M locus is dominant to X. How important is this assumption? Is this compatible with any data (e.g., on feminizing or masculinizing factors in Musca domestica)?

\noindent\paragraph{Response}
\textcolor{blue}{
This is an important assumption.
As discussed in the supplementary of van Doorn \& Kirkpatrick (2007, Nature), partial dominance and incomplete penetrance (partial epistatic dominance) reduce the range parameter values that favour transitions.
However, this reduction in parameter space is smooth and does not preclude transitions, meaning the qualitative conclusions are robust.
Further, there are many systems in which one sex-determining allele is epistatically dominant.
And finally, our Mathematica file allows for arbitrary epistatic dominance coefficients.
We now discuss the empirical evidence for epistatic dominance, our general Mathematica file, and the analysis of van Doorn and Kirkpartick when introducing our model (L117).
}
%\textcolor{red}{This is obviously very important, and is commonly assumed (Kozielska et al. 2010, van Doorn and Kirkpatrick 2007, 2010). There are a large number of other possibilities (given 9 genotypes that determine 2 sexes, ie $2^9$), and it seems like these alternatives could either increase or decrease selection on the invader. van Doorn and Kirkpatrick don't give any motivation (from my quick skim), but Kozielska et al. cite the housefly (Musca domestica), where it seems a single W is epistatically dominant over any number of loci with Y alleles. Is this ``compatible" with our model?}
%
%\textcolor{red}{Our analysis shows that the end state of the system depends on the `epistatic dominance' (term removed from our paper?) of the loci. Could draw this out more for this response. The drift-transitions Genetics paper rests on the difference in dominance, I'm pretty sure they have examples and the majority of new GSD systems are epistatically dominant. Also vD\&K 2010 say that there is evidence that epistatic dominance is common, and explain why a completely recessive should not invade. }

@@

\section{Reviewer 3}

This work (henceforth [SOO]) explores the role of both diploid and haploid selection on the evolution of new sex-chromosomes and or transitions between sex-determination systems. For that purpose, the authors develop a population genetics model of three loci with two alleles in each locus. The first locus X is the ancestral sex-determining locus, the second locus A is the locus under diploid and haploid selection. The third locus M is the novel sex-determining locus.

\noindent\subsection{}
I find the article is not clearly written and could benefit from improving the presentation of the main results. 
%
%\noindent\paragraph{Response}
%\textcolor{blue}{We hope the improvements discussed above (e.g., Response 2.1 and 2.2) and below have increased the clarity of our writing, especially regarding the main results.}
%
%\noindent\subsection{}
I also find the article claims that: (a) the modeling work is novel and (b) the findings are unexpected to be only partially justified.
%
%\noindent\paragraph{Response}
%\textcolor{blue}{As described at length below, we now give more weight to two previous papers (Kozielska et al 2010 and Ubeda et al 2015) and more precisely indicate which of our results are new and unexpected.}
%
%\noindent\subsubsection{}
The works of Kozielska 2014 and Ubeda 2015 model the combined effects of both haploid and diploid selection on the evolution of new sex-chromosomes. The work of [SOO] is thus not novel in considering both haploid and diploid selection on the evolution of new sex-chromosomes. It does generalize previous work by considering sex-specific viability selection and get some nice analytical results.

\noindent\paragraph{Response}
%\textcolor{red}{The abstract and summary still seem to be a little over-selling...}
\textcolor{blue}{We have performed an extensive re-arrangement (e.g., Responses 2.1 and 2.2 and below) to improve the clarity of our writing, especially regarding the main results. 
We believe that our new presentation makes the novel and unexpected features of our results much more apparent. 
In particular, as described at length below, we now give more weight to the two previous papers suggested (Kozielska et al 2010 and Ubeda et al 2015) and more precisely indicate which of our model features and results are new and unexpected in light of these previous results.
To be clear, we do not claim to be the first to include both haploid and diploid selection to a model investigating the evolution of new sex-chromosomes and we do not consider the addition of sex-specific selection or analytical results for previously numerical models to be our most significant contributions (again, see Responses 2.1 and 2.2 and below for more detail on what these are). 
}

@@

\noindent\subsubsection{}
The findings in [SOO] claimed to be surprising (``Surprisingly, we find that neither force (sex ratio selection nor associations with genes that have sex-specific effects) dominates the spread of new sex-determining systems alone.") ... [but this result is rather similar to those in Ubeda et al 2015].

\noindent\paragraph{Response}
\textcolor{blue}{Here, we were referring to the result that $\lambda_{W'}^{XY} = \lambda_{Y'}^{ZW}$ with haploid selection, even when sex-ratio biases evolve to become either stronger or weaker (e.g., $\lambda_{Y'}^{ZW}$ vs. $\lambda_{W'}^{XY}$ with $R<r$ and male meiotic drive) or when sex-ratio biases are either present or absent (e.g., $\lambda_{W'}^{XY}$ vs. $\lambda_{Y'}^{ZW}$ with $r<1/2$, $R=1/2$, and male meiotic drive). The Ubeda et al. (2015) step 2 result shows that associations with a haploid selected locus can favour a new sex-determining allele despite causing biased sex ratios. Nevertheless, it was difficult for us to intuit whether the presence/absence of ancestral sex-ratio bias would have a negligible or overwhelming effect on the evolution of new sex-determining alleles. For example, step 3 of Ubeda et al. (2015) involves the spread of a autosomal meiotic drive suppressor favoured by Fisherian sex-ratio selection and Kozielska et al. (2010) show that ancestral sex-ratio bias can favour new sex determining systems that balance the sex ratio (now discussed in the Introduction, lines 56-66). We found our result, that these forces are exactly equal (with selection weak relative to recombination), such that there is no difference between the propensity for XY-ZW or ZW-XY transitions, to be surprising.
To clarify our meaning we have rephrased (in Author summary): ``Surprisingly, we find the two forces (sex-ratio selection and associations with genes that have sex-specific effects) will often be equally strong, and thus neither dominates the spread of new sex-determining systems alone."}

@@

\noindent\subsubsection{}
\noindent and unexpected (``Even more unexpectedly, we find that, to spread, new sex-determining alleles do not necessarily have to arise in closer linkage with genes that are differentially selected in males and females"), are rather similar to those in Ubeda et al 2015. This is a problem that repeats itself over the paper

\noindent\paragraph{Response}
\textcolor{blue}{The Ubeda et al 2015 results do indeed show that differential selection between male and female diploids is not required for new sex-determining alleles to spread; they show that a new genetic sex-determining allele can spread from ancestral ESD when it is linked to a locus that experiences sex-specific meiotic drive. 
In this sentence, we were attempting to stress that new sex determining alleles can spread when they are \textit{less closely linked or completely unlinked} to loci under differential male/female haploid and/or diploid selection. 
To our knowledge, this has only been shown for the special case of male meiotic drive in an ancestral XY system with $r=0$ and $R=1/2$ with a neo-W, which is driven by sex-ratio selection (Kozielska et al 2010). 
We therefore found the large parameter space allowing invasion with $r<R$, sometimes in the absence of sex-ratio selection (or even in the absence of haploid selection), to be very unexpected.
}
\textcolor{blue}{To communicate this we have rephrased: ``Even more unexpectedly, we find that new sex-determining alleles can spread despite being less closely linked to selected loci under a wide range of conditions." 
We believe that the changes to the presentation of the Introduction and Results should also help to clarify this point further (e.g., paragraph, lines 40-47, and more clearly highlighted conclusions on lines XXX; see Responses 2.1 and 2.2).  }
\textcolor{blue}{We discuss the related results of Kozielska and Ubeda on lines 261 and 350.}

@@

\noindent\subsubsection{}
For example in: Page 2. The authors claim that predominant theories of the evolution of sex-determining systems are two: sexually-antagonistic selection and sex-ratio selection. They ignore other two: inbreeding (Charlesworth 1978) and sex-different drive (Ubeda et al 2015).

\noindent\paragraph{Response}
\textcolor{blue}{
Charlesworth and Charlesworth 1978 (Am Nat and Heredity) discuss the role of inbreeding (and particularly inbreeding depression) in transitions from hermaphroditism to dioecy. 
As noted on lines 468-472 (original submission), we have only considered GSD-GSD transitions and GSD-ESD transitions.
For clarity, we have moved this description into the introduction (lines XXX). 
In addition, where we draw a parallel between ESD and hermaphroditism on lines 451-455 XXX (revision), we now bring up the limitation that our model of ESD does not include inbreeding. }

\textcolor{blue}{
The line referred to used sexually-antagonistic selection as an example of ``selective differences between the sexes". 
We believe the phrase 'selective differences between the sexes' includes sex-differences in drive (Ubeda et al 2015), which just occur during haploid rather than diploid selection (see also Response 3.2.6).
}

\textcolor{blue}{
We have rephrased the line referred to here to clarify our focus on GSD: ``Predominant theories accounting for transitions between GSD systems by selection involve sex-ratio selection or selective differences between the sexes (e.g., sexually antagonistic selection)".
}

### 2018-04-05 LaTeX/main_revised.tex

\end{flushleft}
% Please keep the abstract below 300 words
%currently 308 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We explicitly consider selection on loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection, which is not apparent from previous theory. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
We find that associations with haploid selected loci can drive transitions between sex-determining systems, without requiring selection to act differently in diploid males vs.\ females. 
Further, with haploid selection, transitions between male and female heterogamety can also evolve where linkage with the sex-determining locus is weakened. 
Unexpectedly, we find that the selective strengths exerted by the sex-ratio and associations with selected alleles are equivalent.
This allows sex-ratio biases to increase or decrease with the spread of new sex chromosomes, and implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
%Our results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models extend these theories to include selection upon haploid genotypes (meiotic drive or gametic competition), which causes sex-ratio biases and occurs differently in male and female gametes. 
Surprisingly, we find the two forces (sex-ratio selection and associations with genes that have sex-specific effects) will often be equally strong, and thus neither dominates the spread of new sex-determining systems alone. 
Even more unexpectedly, we find that new sex-determining alleles can spread despite being less closely linked to selected loci under a wide range of conditions. 
Therefore, our models predict loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

%MAJOR MECHANISMS
%(1) sex-ratio selection
Predominant theories accounting for transitions between GSD systems by selection involve sex-ratio selection or selective differences between the sexes (e.g., sexually antagonistic selection)~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
`Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}.
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
Selection therefore typically favours mutants that increase investment in the rarer sex.
Because the sex ratio is directly determined by the sex-determining system, it has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}).

%MAJOR MECHANISMS
%(2) sex-antagonism
Loci experiencing sexual antagonism have been shown to favour the spread of new genetic sex-determining alleles that are closely linked~\cite{vanDoorn:2007eu,vanDoorn:2010hu,Muralidhar2018}. 
%van Doorn and Kirkpatrick~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and Muralidhar and Veller~\cite{Muralidhar2018} have shown that new sex-determining alleles can be favoured if they arise at loci in close linkage with a locus that experiences sexual antagonism. 
Linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y allele, for example. 
Such associations can favour cis-GSD transitions~\cite{vanDoorn:2007eu}, trans-GSD transitions~\cite{vanDoorn:2010hu}, and new partially-masculinizing or partially-feminizing alleles in a population with ESD~\cite{Muralidhar2018}.
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determining locus will develop similar, favourable associations and are expected to hinder the spread of a new sex-determining system~\cite{vanDoorn:2007eu,vanDoorn:2010hu}.

%Tight linkage Model Feature and Result
One novel feature of the models developed here is that, when considering loci in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome), % in the results section: `Tight linkage with the ancestral sex-determining locus'). 
%\textcolor{red}{[vDK2007 does "explicitly" consider this as well (although on pg 15 of their SOM they say they will ignore equilibrium A, as it is stable under a "limited range of biologically relevant parameter combinations"; perhaps haploid selection changes this impression), but only for cis-GSD transitions. They find that if equil B is stable the neo-Y never spreads when r=0. vDK2010 does not "explicitly" calculate pX and pY, which are required to see if their equation 7 is positive or negative, but nowhere do they seem to say that weaker linkage \textit{can't} evolve given r initially small.]}
we explicitly calculate ancestral equilibrium allele frequencies.
This analysis uncovers that, under various forms of selection, the ancestral-X can carry suboptimal alleles for females and/or the ancestral-Y can carry female-beneficial alleles. 
This, in turn, can favour a new ZW sex-determining locus that has weaker linkage with loci under selection, which has not been apparent in previous theory \cite{vanDoorn:2010hu}. 
That is, very tight linkage with the ancestral GSD locus can favour trans-GSD transitions in which associations with selected loci are actually weakened.

@@

%main results
Here, we analytically find the conditions under which new GSD or ESD systems spread with generic linkage between the loci involved (and generic sex-specific haploid and diploid selection). 
Strikingly, we reconcile the results of Kozielska et al. (2010)~\cite{Kozielska:2010vm} and Ubeda et al. (2015)~\cite{Ubeda:2015fx} by showing that new GSD systems commonly spread when they increase \textit{or} decrease linkage with loci that experience haploid selection.
This result is qualitatively distinct from those for diploid selection alone~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and suggests that haploid selection is more likely to promote transitions between sex-determination systems. 
However, we also show that transitions involving haploid selection cannot be simply explained by invoking sex-ratio selection. 
In particular, under a wide range of conditions, we show that haploid selection drives transitions in sex-determining system \textit{equally strongly} in situations where sex-ratio biases increase or decrease (and in situations where sex-ratio biases are present or absent). 
Finally, we show that ESD may not evolve, even if the sex ratio is initially biased by haploid selection, which would not be predicted by previous theories for transitions to ESD~\cite{Charnov:1982wg,Bull:1983vi,West:2009we}. 
Together, in this context, our results suggest that sex ratio selection is not an overwhelming force driving transitions between sex-determining systems.

@@

%HOLES IN CURRENT THEORY
%(1) haploid selection; it exists in many spp
%While both opposing selection in the diploid sexes and sex-ratio selection surely influence the evolution of new sex-determining factors, so too might selection in the haploid phase.
\textcolor{red}{move to discussion somewhere:
Even in animal and plant species that have much larger and more conspicuous diploid phases than haploid phases, many loci experience significant haploid selection through gamete competition and/or meiotic drive~\cite{Mulcahy:1996ha,JOSEPH:2004haa}. }

\textcolor{red}{move to results somewhere:
We use the term `meiotic drive' to refer to the biased (non-Mendelian) segregation of genotypes during gamete production (from one parent only) and the term `gametic competition' to refer to selection upon haploid genotypes within a gamete/gametophyte pool (from many parents); the term `haploid selection' encompasses both processes. }

\textcolor{red}{move to discussion somewhere:
Segregation distortion provides putative evidence of haploid selection and can sometimes be attributed to meiotic drive and/or gametic competition~\cite{Lalanne2004,Fishman2005,Leppala2008,Leppala2013,Didion2015,Didion2016}.
Where it has been characterized, individual cases of meiotic drive are generally sex-limited, exclusively affecting male or female gametes~\cite{Ubeda:2005gw,Lindholm:2016cw}.
Gametic competition is also typically sex-limited, occurring primarily among male gametes, because there are typically many more pollen/sperm than required for fertilization.
Gametic competition may be particularly common in plants, in which 60-70\% of all genes are expressed in the male gametophyte, and these genes exhibit stronger signatures of selection than randomly-chosen genes~\cite{Borg:2009jpa,Arunkumar:2013iq,Gossmann:2014dua}.
In addition, artificial selection pressures applied to male gametophytes are known to cause a response to selection (e.g.,~\cite{Hormaza:1996cv,Ravikumar:2003uo,Hedhly:2004iv,Clarke:2004ir}). 
A smaller proportion of genes are thought to be expressed and selected during competition in animal sperm, although precise estimates are uncertain~\cite{Zheng:2001fi,JOSEPH:2004haa,Vibranovski:2010et}. 
Nevertheless, recent studies have demonstrated that sperm competition in animals, even within a single ejaculate, can alter haploid allele frequencies and increase offspring fitness~\cite{Immler:2014im,Alavioon2017}.
}

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
%currently 308 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We consider the dynamics of loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection, which is not expected from previous theory. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
Haploid selection can drive transitions between sex-determining systems without requiring selection to act differently in diploid males vs.\ females. 
With haploid selection, we find that transitions between male and female heterogamety can evolve where linkage with the sex-determining locus is either strengthened or weakened. 
Furthermore, we find that the selective strengths exerted by the sex-ratio and associations with selected alleles are often equivalent. 
This allows sex-ratio biases to increase or decrease with the spread of new sex chromosomes, and implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
%Our results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models include selection upon haploid genotypes (meiotic drive or gametic competition), which causes sex-ratio biases and occurs differently in male and female gametes. 
Surprisingly, we find the two forces (sex-ratio selection and associations with genes that have sex-specific effects) will often be equally strong, and thus neither dominates the spread of new sex-determining systems alone. 
Even more unexpectedly, we find that new sex-determining alleles can spread despite being less closely linked to selected loci. 
Therefore, our models predict loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

%MAJOR MECHANISMS
%(1) sex-ratio selection
%Predominant theories accounting for transitions between GSD systems by selection involve sex-ratio selection or selective differences between the sexes (e.g., sexually antagonistic selection)~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
It has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination because the sex ratio is directly determined by the sex-determining system (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
Classic `Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}. 
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
Selection therefore typically favours mutants that increase investment in the rarer sex.
%Because the sex ratio is directly determined by the sex-determining system, it has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}).

%MAJOR MECHANISMS
%(2) sex-antagonism
The evolution of sex determination is also thought to be strongly influenced by differences in selection between the sexes ~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}.
For example, loci experiencing sexual antagonism have been shown to favour the spread of new genetic sex-determining alleles that are closely linked~\cite{vanDoorn:2007eu,vanDoorn:2010hu,Muralidhar2018}. 
%van Doorn and Kirkpatrick~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and Muralidhar and Veller~\cite{Muralidhar2018} have shown that new sex-determining alleles can be favoured if they arise at loci in close linkage with a locus that experiences sexual antagonism. 
Linkage allows a stronger favourable association to build up between a male-beneficial allele, and a neo-Y allele, for example. 
Such associations can favour cis-GSD transitions~\cite{vanDoorn:2007eu}, trans-GSD transitions~\cite{vanDoorn:2010hu}, and new partially-masculinizing or partially-feminizing alleles in a population with ESD~\cite{Muralidhar2018}.
However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determining locus should develop similar, favourable associations and are expected to hinder the spread of a new sex-determining system~\cite{vanDoorn:2007eu,vanDoorn:2010hu}.

%Tight linkage Model Feature and Result
One novel feature of the models developed here is that we explicitly consider the maintenance of genetic variation around the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome). %during trans-GSD transitions
%, when considering loci in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome), 
%\textcolor{red}{[vDK2007 does "explicitly" consider this as well (although on pg 15 of their SOM they say they will ignore equilibrium A, as it is stable under a "limited range of biologically relevant parameter combinations"; perhaps haploid selection changes this impression), but only for cis-GSD transitions. They find that if equil B is stable the neo-Y never spreads when r=0. vDK2010 does not "explicitly" calculate pX and pY, which are required to see if their equation 7 is positive or negative, but nowhere do they seem to say that weaker linkage \textit{can't} evolve given r initially small.]}
%we explicitly calculate ancestral polymorphisms. 
Our analysis reveals that, under various forms of selection, the ancestral-X can carry suboptimal alleles for females and/or the ancestral-Y can carry female-beneficial alleles. 
This, in turn, can favour a new ZW sex-determining locus that has weaker linkage with loci under selection, which has not been apparent in previous theory \cite{vanDoorn:2010hu}. 
That is, we show that selected loci in very tight linkage with the ancestral GSD locus can favour trans-GSD transitions in which associations are actually weakened.

@@

%main results
Here, we analytically find the conditions under which new GSD or ESD systems spread with generic linkage between the loci involved (and generic sex-specific haploid and diploid selection). 
Strikingly, we reconcile the results of Kozielska et al. (2010)~\cite{Kozielska:2010vm} and Ubeda et al. (2015)~\cite{Ubeda:2015fx} by showing that new GSD systems commonly spread when they increase \textit{or} decrease linkage with loci that experience haploid selection.
This result is qualitatively distinct from those for diploid selection alone~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and suggests that haploid selection is more likely to promote transitions between sex-determination systems. 
However, we also show that transitions involving haploid selection cannot be simply explained by invoking sex-ratio selection. 
In particular, under a wide range of conditions, we show that transitions in sex-determining system are favoured \textit{equally strongly} in situations where sex-ratio biases increase or decrease (and in situations where sex-ratio biases are present or absent). 
Finally, we show that ESD may not evolve, even if the sex ratio is initially biased by haploid selection, which would not be predicted by previous theories for transitions to ESD~\cite{Charnov:1982wg,Bull:1983vi,West:2009we}. 
Together, our results suggest that sex ratio selection is not an overwhelming force driving transitions between sex-determining systems.

@@

%HOLES IN CURRENT THEORY
%(1) haploid selection; it exists in many spp
%While both opposing selection in the diploid sexes and sex-ratio selection surely influence the evolution of new sex-determining factors, so too might selection in the haploid phase.
